---
title: "Lipoproteins and cholesterol efflux"
subtitle: "RSTR vs. LTBI"
author: "Kim Dill-McFarland, kadm@uw.edu"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
date: "version `r format(Sys.time(), '%B %d, %Y')`"
editor_options: 
  chunk_output_type: console
---
# Setup
Load packages

```{r message=FALSE, warning=FALSE}
# Data manipulation and figures
library(tidyverse)
library(readxl)
library(BIGpicture)
library(ggrepel)
library(patchwork)
# Linear modeling
library(kimma)
#Print pretty tables to Rmd
library(knitr)
library(kableExtra)
options(knitr.kable.NA = '')
```

Set seed

```{r}
set.seed(589)
```

# Load data

From Tomas Vaisar

```{r}
dat <- list()
dat$targets <- read_csv("../data_refs/2021.09.08RSTR_Hawn_metadata.csv") %>% 
  distinct(FULLIDNO, RS_SUB_ACCESSION_NO, Sample_Group, M0_KCVSEX, KCHCA_AGE_YR_CURRENT,
           avgBMI, RISK_SCORE) %>% 
  drop_na(RS_SUB_ACCESSION_NO) %>% 
  arrange(FULLIDNO)

dat$E <- read_excel("data_raw/Study191_DMA_Efflux_ForKim211012.xlsx", 
                    sheet="DMA.data") %>% 
  select(AssayID, SUB_ACCESSION_NO, tHDL:xlHDL.sz) %>% 
  full_join(read_excel("data_raw/Study191_DMA_Efflux_ForKim211012.xlsx", 
                       sheet="Efflux.data"), by = "AssayID") %>% 
  select(-Group) %>% 
  left_join(distinct(dat$targets, RS_SUB_ACCESSION_NO, FULLIDNO),
            by=c("SUB_ACCESSION_NO"="RS_SUB_ACCESSION_NO")) %>% 
  select(FULLIDNO, tHDL:ABCA1.spec) %>% 
  arrange(FULLIDNO) %>% 
  column_to_rownames("FULLIDNO") %>% 
  t() %>% as.data.frame()

dat$targets <- dat$targets %>% 
  filter(FULLIDNO %in% colnames(dat$E)) %>% 
  distinct(FULLIDNO, Sample_Group, M0_KCVSEX, KCHCA_AGE_YR_CURRENT,
           avgBMI, RISK_SCORE)
```

```{r}
#Check one measure per individual
length(unique(dat$targets$FULLIDNO)) == ncol(dat$E)
```

## Kinship

```{r echo=FALSE, message=FALSE}
kin <- read_csv("../data_refs/kinship_Hawn_all.csv")
overlap <- intersect(kin$rowname, dat$targets$FULLIDNO)

kin <- kin %>% 
  filter(rowname %in% overlap) %>% 
  column_to_rownames() %>% 
  select(all_of(overlap)) %>% 
  as.matrix()
```

## Explore data

Nothing stands out as particularly non-normal.

```{r echo=FALSE, message=FALSE}
dat$E %>% 
  rownames_to_column() %>%
  pivot_longer(-rowname, names_to = "FULLIDNO") %>% 
  left_join(dat$targets) %>% 
  ggplot(aes(x=value)) +
  geom_histogram(bins = 40) +
  theme_classic() +
  facet_wrap(~rowname, scales = "free") +
  labs(x="")
```

Differences between groups are not large.

```{r echo=FALSE, message=FALSE, fig.height=10}
dat$E %>% 
  rownames_to_column() %>%
  pivot_longer(-rowname, names_to = "FULLIDNO") %>% 
  left_join(dat$targets) %>% 
  ggplot(aes(x=Sample_Group, y=value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width=0.2, height=0) +
  theme_classic() +
  facet_wrap(~rowname, scales = "free") +
  labs(x="",y="")
```

# Linear modeling
## CEC

```{r}
dat.cec <- dat
dat.cec$E <- as.data.frame(dat.cec$E) %>% 
  rownames_to_column() %>% 
  filter(rowname %in% c("J774.ind","J774.basal","J774.abca1",
                        "ABCA1.ind","ABCA1.basal","ABCA1.spec")) %>% 
  column_to_rownames()
```

```{r}
model.kin <- kmFit(dat.cec, kin, model="~Sample_Group+(1|FULLIDNO)", 
                   run_lmerel = TRUE, 
                   patientID = "FULLIDNO", libraryID = "FULLIDNO",
                   metrics=TRUE)
```

### Clinical covariates

Sex 

```{r}
model.kin.sex <- kmFit(dat.cec, kin,
                       model="~Sample_Group+M0_KCVSEX+(1|FULLIDNO)", 
                       run_lmerel = TRUE, 
                       patientID = "FULLIDNO", libraryID = "FULLIDNO",
                       metrics=TRUE)
```

Age

```{r}
model.kin.age <- kmFit(dat.cec, kin,
                       model="~Sample_Group+KCHCA_AGE_YR_CURRENT+(1|FULLIDNO)", 
                       run_lmerel = TRUE, 
                       patientID = "FULLIDNO", libraryID = "FULLIDNO",
                       metrics = TRUE)
```

BMI

```{r}
model.kin.bmi <- kmFit(dat.cec, kin,
                       model="~Sample_Group+avgBMI+(1|FULLIDNO)", 
                       run_lmerel = TRUE, 
                       patientID = "FULLIDNO", libraryID = "FULLIDNO",
                       metrics=TRUE)
```

Compare fits

```{r}
plot_fit2(model.kin, model.kin.sex, x="lmerel", y="lmerel",
          metrics = c("AIC"))[[1]] +
  labs(title = "+/- sex") +
  plot_fit2(model.kin, model.kin.age, x="lmerel", y="lmerel",
            metrics = c("AIC"))[[1]] +
  labs(title = "+/- age") +
  plot_fit2(model.kin, model.kin.bmi, x="lmerel", y="lmerel",
            metrics = c("AIC"))[[1]] +
  labs(title = "+/- BMI")
```

BMI improves model fit for all CEC measures while age and sex tend to slightly worsen model fit, though not to a significant degree (\delta AIC < ).

### Full model

`~ RSTR + BMI + age + sex + kinship`

```{r}
model.all <- kmFit(dat.cec, kin,
                   model="~Sample_Group+avgBMI+KCHCA_AGE_YR_CURRENT+M0_KCVSEX+(1|FULLIDNO)",
                   run_lmerel = TRUE, 
                   patientID = "FULLIDNO", libraryID = "FULLIDNO",
                   metrics = TRUE)

summarise_kmFit(model.all$lmerel, p.cutoff=c(0.01,0.05,0.1))

plot_fit2(model.kin, model.all, x="lmerel", y="lmerel",
          metrics = c("AIC")) 
```

## HDL concentrations and sizes

```{r}
dat.con <- dat

dat.con$E <- as.data.frame(dat.con$E) %>% 
  rownames_to_column() %>% 
  filter(rowname %in% c("xsHDL","sHDL","mHDL","mlHDL",
                        "lHDL","xlHDL","tHDL")) %>% 
  column_to_rownames()

dat.sz <- dat

dat.sz$E <- as.data.frame(dat.sz$E) %>% 
  rownames_to_column() %>% 
  filter(rowname %in% c("xsHDL.sz","sHDL.sz","mHDL.sz","mlHDL.sz",
                        "lHDL.sz","xlHDL.sz")) %>% 
  column_to_rownames()
```

```{r}
model.kin2a <- kmFit(dat.con, kin, model="~Sample_Group+(1|FULLIDNO)", 
                   run_lmerel = TRUE, 
                   patientID = "FULLIDNO", libraryID = "FULLIDNO",
                   metrics=TRUE)
model.kin2b <- kmFit(dat.sz, kin, model="~Sample_Group+(1|FULLIDNO)", 
                   run_lmerel = TRUE, 
                   patientID = "FULLIDNO", libraryID = "FULLIDNO",
                   metrics=TRUE)
```

### Clinical covariates

Sex 

```{r}
model.kin.sex2a <- kmFit(dat.con, kin,
                       model="~Sample_Group+M0_KCVSEX+(1|FULLIDNO)", 
                       run_lmerel = TRUE, 
                       patientID = "FULLIDNO", libraryID = "FULLIDNO",
                       metrics=TRUE)
model.kin.sex2b <- kmFit(dat.sz, kin,
                       model="~Sample_Group+M0_KCVSEX+(1|FULLIDNO)", 
                       run_lmerel = TRUE, 
                       patientID = "FULLIDNO", libraryID = "FULLIDNO",
                       metrics=TRUE)
```

Age

```{r}
model.kin.age2a <- kmFit(dat.con, kin,
                       model="~Sample_Group+KCHCA_AGE_YR_CURRENT+(1|FULLIDNO)", 
                       run_lmerel = TRUE, 
                       patientID = "FULLIDNO", libraryID = "FULLIDNO",
                       metrics = TRUE)
model.kin.age2b <- kmFit(dat.sz, kin,
                       model="~Sample_Group+KCHCA_AGE_YR_CURRENT+(1|FULLIDNO)", 
                       run_lmerel = TRUE, 
                       patientID = "FULLIDNO", libraryID = "FULLIDNO",
                       metrics = TRUE)
```

BMI

```{r}
model.kin.bmi2a <- kmFit(dat.con, kin,
                       model="~Sample_Group+avgBMI+(1|FULLIDNO)", 
                       run_lmerel = TRUE, 
                       patientID = "FULLIDNO", libraryID = "FULLIDNO",
                       metrics=TRUE)
model.kin.bmi2b <- kmFit(dat.sz, kin,
                       model="~Sample_Group+avgBMI+(1|FULLIDNO)", 
                       run_lmerel = TRUE, 
                       patientID = "FULLIDNO", libraryID = "FULLIDNO",
                       metrics=TRUE)
```

Compare fits for HDL concentrations

```{r}
plot_fit2(model.kin2a, model.kin.sex2a, x="lmerel", y="lmerel",
                metrics = c("AIC"))[[1]] +
  labs(title = "+/- sex") +
  plot_fit2(model.kin2a, model.kin.age2a, x="lmerel", y="lmerel",
                metrics = c("AIC"))[[1]] +
  labs(title = "+/- age") +
  plot_fit2(model.kin2a, model.kin.bmi2a, x="lmerel", y="lmerel",
                metrics = c("AIC"))[[1]] +
  labs(title = "+/- BMI")
```

Compare fits for HDL sizes

```{r}
plot_fit2(model.kin2b, model.kin.sex2b, x="lmerel", y="lmerel",
                metrics = c("AIC"))[[1]] +
  labs(title = "+/- sex") +
  plot_fit2(model.kin2b, model.kin.age2b, x="lmerel", y="lmerel",
                metrics = c("AIC"))[[1]] +
  labs(title = "+/- age") +
  plot_fit2(model.kin2b, model.kin.bmi2b, x="lmerel", y="lmerel",
                metrics = c("AIC"))[[1]] +
  labs(title = "+/- BMI")
```

### Full model

`HDL concentration ~ RSTR + BMI + age + sex + kinship`

```{r}
model.all2a <- kmFit(dat.con, kin,
                   model="~Sample_Group+avgBMI+KCHCA_AGE_YR_CURRENT+M0_KCVSEX+(1|FULLIDNO)",
                   run_lmerel = TRUE, 
                   patientID = "FULLIDNO", libraryID = "FULLIDNO",
                   metrics = TRUE)

summarise_kmFit(model.all2a$lmerel, fdr.cutoff=c(0.01,0.05,0.1))

plot_fit2(model.kin2a, model.all2a, x="lmerel", y="lmerel",
          metrics = c("AIC")) 
```

`HDL size ~ RSTR + BMI + age + sex + kinship`

```{r}
model.all2b <- kmFit(dat.sz, kin,
                   model="~Sample_Group+avgBMI+KCHCA_AGE_YR_CURRENT+M0_KCVSEX+(1|FULLIDNO)",
                   run_lmerel = TRUE, 
                   patientID = "FULLIDNO", libraryID = "FULLIDNO",
                   metrics = TRUE)

summarise_kmFit(model.all2b$lmerel, p.cutoff=c(0.01,0.05,0.1))

plot_fit2(model.kin2b, model.all2b, x="lmerel", y="lmerel",
          metrics = c("AIC")) 
```

## HDL proportions

Calculate HDL measures as a percentage of total.

```{r}
dat.conp <- dat.con

dat.conp$E <- as.data.frame(dat.conp$E) %>% 
  t() %>% as.data.frame() %>% 
  mutate(across(xsHDL:xlHDL, ~./tHDL, .names = "{.col}_pct")) %>% 
  select(contains("_pct")) %>% 
  t()
```

```{r}
model.kin3 <- kmFit(dat.conp, kin, model="~Sample_Group+(1|FULLIDNO)", 
                   run_lmerel = TRUE, 
                   patientID = "FULLIDNO", libraryID = "FULLIDNO",
                   metrics=TRUE)
```

### Clinical covariates

Sex 

```{r}
model.kin.sex3 <- kmFit(dat.conp, kin,
                       model="~Sample_Group+M0_KCVSEX+(1|FULLIDNO)", 
                       run_lmerel = TRUE, 
                       patientID = "FULLIDNO", libraryID = "FULLIDNO",
                       metrics=TRUE)
```

Age

```{r}
model.kin.age3 <- kmFit(dat.conp, kin,
                       model="~Sample_Group+KCHCA_AGE_YR_CURRENT+(1|FULLIDNO)", 
                       run_lmerel = TRUE, 
                       patientID = "FULLIDNO", libraryID = "FULLIDNO",
                       metrics = TRUE)
```

BMI

```{r}
model.kin.bmi3 <- kmFit(dat.conp, kin,
                       model="~Sample_Group+avgBMI+(1|FULLIDNO)", 
                       run_lmerel = TRUE, 
                       patientID = "FULLIDNO", libraryID = "FULLIDNO",
                       metrics=TRUE)
```

Compare fits for HDL proportions

```{r}
plot_fit2(model.kin3, model.kin.sex3, x="lmerel", y="lmerel",
                metrics = c("AIC"))[[1]] +
  labs(title = "+/- sex") +
  plot_fit2(model.kin3, model.kin.age3, x="lmerel", y="lmerel",
                metrics = c("AIC"))[[1]] +
  labs(title = "+/- age") +
  plot_fit2(model.kin3, model.kin.bmi3, x="lmerel", y="lmerel",
                metrics = c("AIC"))[[1]] +
  labs(title = "+/- BMI")
```

### Full model

`HDL proportion ~ RSTR + BMI + age + sex + kinship`

```{r}
model.all3 <- kmFit(dat.conp, kin,
                   model="~Sample_Group+avgBMI+KCHCA_AGE_YR_CURRENT+M0_KCVSEX+(1|FULLIDNO)",
                   run_lmerel = TRUE, 
                   patientID = "FULLIDNO", libraryID = "FULLIDNO",
                   metrics = TRUE)

summarise_kmFit(model.all3$lmerel, fdr.cutoff=c(0.01,0.05,0.1))

plot_fit2(model.kin3, model.all3, x="lmerel", y="lmerel",
          metrics = c("AIC")) 
```

## CEC ~ HDL

Compare HDL concentrations and proportions to CEC.

```{r}
dat2 <- dat
dat2$E <- as.data.frame(dat2$E) %>% 
  rownames_to_column() %>% 
  filter(rowname %in% c("J774.ind","ABCA1.spec")) %>% 
  column_to_rownames()

#Move HDL to x variables
dat2$targets <- dat2$targets %>% 
  #concentrations and sizes
  full_join(as.data.frame(t(dat.con$E)) %>% 
              rownames_to_column("FULLIDNO"),
            by="FULLIDNO") %>% 
  #sizes
  # full_join(as.data.frame(t(dat.sz$E)) %>% 
  #             rownames_to_column("FULLIDNO"),
  #           by="FULLIDNO") %>% 
  #proportions
  full_join(as.data.frame(t(dat.conp$E)) %>% 
              rownames_to_column("FULLIDNO"),
            by="FULLIDNO")
```

### Full model

`~ HDL measure + BMI + age + sex + kinship`

*Note*: The interaction of RSTR:HDL was not significant for any model (FDR > 0.5).

```{r message=FALSE}
model.cec.hdl <- data.frame()

for(x in colnames(dat2$targets[7:ncol(dat2$targets)])){
  print(x)
  m.temp <- paste0("~",x,
                   "+avgBMI+KCHCA_AGE_YR_CURRENT+M0_KCVSEX+(1|FULLIDNO)")
  
  model.temp <- kmFit(dat2, kin, 
                      model=m.temp, 
                      run_lmerel = TRUE, 
                      patientID = "FULLIDNO", libraryID = "FULLIDNO",
                      metrics=FALSE)
  model.cec.hdl <- model.temp$lmerel %>% 
    mutate(model = x) %>% 
    bind_rows(model.cec.hdl)
}

#Recalc FDR for multiple x tests
model.cec.hdl.fdr <- model.cec.hdl %>% 
  mutate(model_group = case_when(grepl("_pct$",model)~"pct",
                                 grepl(".sz$",model)~"sz",
                                 TRUE~"HDL")) %>% 
  mutate(variable_group = case_when(
    # grepl(":",variable)~"Sample_Group:HDL",
    grepl("_pct$",variable)~"pct",
    grepl(".sz$",variable)~"sz",
    variable %in% c("avgBMI","KCHCA_AGE_YR_CURRENT",
                    "M0_KCVSEX","(1 | FULLIDNO)","Sample_Group")~variable,
    TRUE~"HDL")) %>% 
  group_by(gene,model_group,variable_group) %>% 
  mutate(FDR=p.adjust(pval, method="BH")) %>% 
  ungroup()

# count(model.cec.hdl.fdr,gene,model_group,variable_group) %>% View
```

```{r}
summarise_kmFit(model.cec.hdl.fdr, p.cutoff=c(0.01,0.05,0.1))
```

# Significant results

Combine results and save.

```{r message=FALSE}
fdr <- model.all$lmerel %>% 
  mutate(group="CEC") %>% 
  bind_rows(model.all2a$lmerel %>% 
              mutate(group="HDL")) %>% 
  bind_rows(model.all2b$lmerel %>% 
              mutate(group="HDL.sz")) %>% 
  bind_rows(model.all3$lmerel %>% 
              mutate(group="HDL.pct"))

write_csv(fdr, file="results/lipoprot_efflux_model_results.csv")

write_csv(model.cec.hdl.fdr,
          file="results/lipoprot_efflux-HDL_model_results.csv")
```

### CEC

```{r}
dat3 <- dat$E %>% 
  t() %>% as.data.frame() %>% 
  rownames_to_column("FULLIDNO") %>% 
  filter(FULLIDNO %in% rownames(kin)) %>% 
  pivot_longer(-FULLIDNO, names_to = "gene") %>% 
  bind_rows(dat.conp$E %>% 
              t() %>% as.data.frame() %>% 
              rownames_to_column("FULLIDNO") %>% 
              filter(FULLIDNO %in% rownames(kin)) %>% 
              pivot_longer(-FULLIDNO, names_to = "gene")) %>% 
  left_join(dat$targets, by = "FULLIDNO") %>% 
  left_join(fdr, by = "gene") %>% 
  filter(!gene %in% c("ssHDL","msuHDL"))

dat3 %>% 
  filter(variable == "Sample_Group" & pval < 0.2 & group == "CEC") %>% 
  mutate(facet.lab = paste0(gene, "\nP = ", round(pval, digits=3), "\n",
                            "FDR = ", round(FDR, digits=3))) %>% 
  mutate(age_group = cut(KCHCA_AGE_YR_CURRENT, seq(10, 70, 20))) %>% 
  
  ggplot(aes(x=Sample_Group, y=value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(color=age_group), width=0.2, height=0) +
  theme_classic() +
  facet_wrap(~facet.lab, scales = "free") +
  labs(x="",y="",color="Age (yrs)")
```

### HDL concentration, proportion, size

No significant results.

```{r}
dat3 %>% 
  filter(variable == "Sample_Group" & FDR < 0.2 & 
           group %in% c("HDL","HDL.sz","HDL.pct"))
```

### CEC ~ HDL

```{r}
# model.cec.hdl.fdr %>%
#   filter(variable_group %in% c("pct","HDL","sz","Sample_Group:HDL") & FDR < 0.2)

dat4 <- dat$E %>% 
  t() %>% as.data.frame() %>% 
  rownames_to_column("FULLIDNO") %>% 
  filter(FULLIDNO %in% rownames(kin)) %>% 
  select(FULLIDNO:xlHDL.sz) %>% 
  full_join(dat.conp$E %>% 
              t() %>% as.data.frame() %>% 
              rownames_to_column("FULLIDNO") %>% 
              filter(FULLIDNO %in% rownames(kin))) %>% 
  pivot_longer(-FULLIDNO, names_to = "model", values_to = "x") %>% 
  left_join(dat$E %>% 
              t() %>% as.data.frame() %>% 
              rownames_to_column("FULLIDNO") %>% 
              filter(FULLIDNO %in% rownames(kin)) %>% 
              select(FULLIDNO, J774.ind, ABCA1.spec)) %>% 
  pivot_longer(J774.ind:ABCA1.spec, names_to = "gene", values_to = "y") %>%
  inner_join(model.cec.hdl.fdr %>% 
             filter(variable_group %in% c("pct","HDL"))) %>% 
  filter(FDR < 0.2) %>% 
  left_join(dat$targets %>% select(FULLIDNO,Sample_Group))

p1 <- dat4 %>% 
  mutate(facet.lab = paste0(gene, "\nP = ", round(pval, digits=3), "\n",
                            "FDR = ", round(FDR, digits=3))) %>% 

  ggplot(aes(x=x, y=y)) +
  geom_point(aes(color=Sample_Group)) +
  geom_smooth(se=FALSE, method="lm", formula="y~x") +
  theme_classic() +
  facet_wrap(model~facet.lab, scales = "free") +
  labs(x="",y="J774 CEC (%)",color="")

p1
ggsave("CEC_HDL.png",p1,width=6, height=3)
```

# Check age

```{r}
dat3 %>% 
  distinct(FULLIDNO, Sample_Group, KCHCA_AGE_YR_CURRENT) %>% 
  ggplot(aes(x=Sample_Group, y=KCHCA_AGE_YR_CURRENT)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width=0.2, height=0) +
  theme_classic() +
  labs(x="")

dat3 %>% 
  distinct(FULLIDNO, Sample_Group, KCHCA_AGE_YR_CURRENT) %>% 
  t.test(KCHCA_AGE_YR_CURRENT~Sample_Group, data=.)
```

# R session

```{r}
sessionInfo()
```

***
